\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2015-05-21 doc/USER_MANUAL/06_fault_sources.tex

Faults need to be handled in a special way during mesh generation.
A fault surface must lie at the interface between elements (the mesh
must honor the fault surfaces). Moreover, a fault is made of two surfaces
in contact. Each of these two surfaces needs a separate set of nodes.
This approach is known as \textquotedbl{}split nodes\textquotedbl{}.
Currently faults can only be run with \texttt{xdecompose\_mesh} and
\texttt{CUBIT}. \texttt{xmeshfem3D} is not yet ready to handle faults.\\
To facilitate the mesh generation with split nodes in CUBIT, we need
to separate the two fault surfaces by a small distance, effectively
creating a tiny opening of the fault (Figure \ref{fig:examples.splitnodes},
\ref{fig:examples.splitnodes-surfacetrace}). Note that only the interior
of the fault must be opened, its edges must remain closed (except
the edge on the free surface). The fault is automatically closed later
by SPECFEM3D.\\

Here is an example CUBIT script to generate a mesh with split nodes
for a buried vertical strike-slip fault:
{\footnotesize
\begin{verbatim}
 reset
 brick x 10 y 10 z 10
 webcut volume all with plane xplane
 webcut volume all with plane yplane
 webcut volume all with plane xplane offset 3
 webcut volume all with plane zplane offset 3
 webcut volume all with plane zplane offset -3
 imprint all
 merge all
 unmerge surf 160
 mesh vol all
 set node constraint off
 node in surf 168 move X 0 Y 0.01 Z 0
 node in surf 160 move X 0 Y -0.01 Z 0
\end{verbatim}
}
%
\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[scale=0.3]{figures/faultmesh.jpg} \\
\includegraphics[scale=0.3]{figures/surf168.jpg}
\includegraphics[scale=0.3]{figures/splitnodes.jpg}
\par\end{centering}

\caption{Screenshots of the CUBIT example for the embedded fault with split
nodes: The entire mesh is decomposed into several volumes bounding
the fault (top), the fault surface 168 is shown as orange squares
(middle) and split nodes of the fault shown as orange dots(bottom).
Note that only interior nodes of the fault are split while those on
the edges of the fault surface are touching each other.}
\label{fig:examples.splitnodes}
\end{figure}

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[scale=0.55]{figures/splitnodes_surfacetrace.pdf} \\

\par\end{centering}

\caption{Screenshots of a CUBIT example showing split nodes for a fault reaching
the surface. Surface trace of the fault is shown in orange. Note that
the edges of the fault are not split while the interior nodes are
offset by a small distance on either side of the fault}

\label{fig:examples.splitnodes-surfacetrace}
\end{figure}

The CUBIT scripts ({*}.jou and {*}.py) in the directory \texttt{EXAMPLES}
generate more complicated meshes. The {*}.py files are Python scripts
that execute CUBIT commands and use the CUBIT-python interface for
SPECFEM3D (see next section). The Python language allows to define
and manipulate variables to parameterize the mesh. Alternatively,
the Python script can call a CUBIT journal file ({*}.jou), which looks
like the example above. Variables can be defined and manipulated there
using the $\mathtt{APREPRO}$ language built in CUBIT.\\

Note that you should avoid gaps in the list of indices of mesh objects
with the following CUBIT command:
{\footnotesize
\begin{verbatim}
compress ids hex face edge node
\end{verbatim}
}
(otherwise you will get a segmentation fault during domain decomposition)

@@

The directories \texttt{EXAMPLES/*/post} contain additional Matlab
scripts to generate figures specific to each example.

>>>>

Faults need to be handled in a special way during mesh generation.
A fault surface must lie at the interface between elements (the mesh
must honor the fault surfaces). Moreover, a fault is made of two surfaces
in contact. Each of these two surfaces needs a separate set of nodes.
This approach is known as \textquotedbl{}split nodes\textquotedbl{}.
Currently faults can only be run with \texttt{xdecompose\_mesh} and
\texttt{CUBIT}. \texttt{xmeshfem3D} is not yet ready to handle faults.\\

To facilitate the mesh generation with split nodes in CUBIT, we need
to separate the two fault surfaces by a small distance, effectively
creating a tiny opening of the fault (Figure \ref{fig:examples.splitnodes},
\ref{fig:examples.splitnodes-surfacetrace}). Note that only the interior
of the fault must be opened, its edges must remain closed (except
the edge on the free surface). The fault is automatically closed later
by SPECFEM3D.\\

\noindent
Here is an example CUBIT script to generate a mesh with split nodes
for a buried vertical strike-slip fault:

{\footnotesize
\begin{verbatim}
 reset
 brick x 10 y 10 z 10
 webcut volume all with plane xplane
 webcut volume all with plane yplane
 webcut volume all with plane xplane offset 3
 webcut volume all with plane zplane offset 3
 webcut volume all with plane zplane offset -3
 imprint all
 merge all
 unmerge surf 160
 mesh vol all
 set node constraint off
 node in surf 168 move X 0 Y 0.01 Z 0
 node in surf 160 move X 0 Y -0.01 Z 0
\end{verbatim}
}

\begin{figure}[htbp]
\begin{centering}
\includegraphics[scale=0.3]{figures/faultmesh.jpg} \\
\includegraphics[scale=0.3]{figures/surf168.jpg}
\includegraphics[scale=0.3]{figures/splitnodes.jpg}
\par
\end{centering}
\caption{Screenshots of the CUBIT example for the embedded fault with split
nodes: The entire mesh is decomposed into several volumes bounding
the fault (top), the fault surface 168 is shown as orange squares
(middle) and split nodes of the fault shown as orange dots(bottom).
Note that only interior nodes of the fault are split while those on
the edges of the fault surface are touching each other.}
\label{fig:examples.splitnodes}
\end{figure}

\begin{figure}[htbp]
\begin{centering}
\includegraphics[scale=0.55]{figures/splitnodes_surfacetrace.pdf}
\par
\end{centering}
\caption{Screenshots of a CUBIT example showing split nodes for a fault reaching
the surface. Surface trace of the fault is shown in orange. Note that
the edges of the fault are not split while the interior nodes are
offset by a small distance on either side of the fault}
\label{fig:examples.splitnodes-surfacetrace}
\end{figure}

\noindent
The CUBIT scripts ({*}.jou and {*}.py) in the directory \texttt{EXAMPLES}
generate more complicated meshes. The {*}.py files are Python scripts
that execute CUBIT commands and use the CUBIT-python interface for
SPECFEM3D (see next section). The Python language allows to define
and manipulate variables to parameterize the mesh. Alternatively,
the Python script can call a CUBIT journal file ({*}.jou), which looks
like the example above. Variables can be defined and manipulated there
using the $\mathtt{APREPRO}$ language built in CUBIT.\\

\newpage
\noindent
Note that you should avoid gaps in the list of indices of mesh objects
with the following CUBIT command:

{\footnotesize
\begin{verbatim}
compress ids hex face edge node
\end{verbatim}
}
\noindent
(otherwise you will get a segmentation fault during domain decomposition)

@@

\noindent
The directories \texttt{EXAMPLES/*/post} contain additional Matlab
scripts to generate figures specific to each example.

### 2015-05-21 doc/USER_MANUAL/07_adjoint_simulations.tex

\begin{enumerate}
\item First, run a regular forward simlation (\texttt{SIMULATION\_TYPE =
1} and \texttt{SAVE\_FORWARD = .false.}). You can automatically set
these two variables using the \texttt{\small utils/change\_simulation\_type.pl}{\small{}
script:}{\small \par}

\begin{lyxcode}
utils/change\_simulation\_type.pl~-f~
\end{lyxcode}

@@

\begin{lyxcode}
xcreate\_adjsrc\_traveltime~t1~t2~ifile{[}0-5{]}~E/N/Z-ascii-files~{[}baz{]}
\end{lyxcode}

where \texttt{t1} and \texttt{t2} are the start and end time of the
portion you are interested in, \texttt{ifile} denotes the component
of the seismograms to be used (0 for all three components, 1 for East,
2 for North, and 3 for vertical, 4 for transverse, and 5 for radial
component), \texttt{E/N/Z-ascii-files} indicate the three-component
displacement seismograms in the right order, and \texttt{baz} is the
back-azimuth of the station. Note that \texttt{baz} is only supplied
when \texttt{ifile} = 4 or 5.

\item Similarly, a sample program to compute adjoint sources for amplitude
finite-frequency kernels may be found in \texttt{utils/adjoint\_sources/amplitude}
and used in the same way as described for traveltime measurements

\begin{lyxcode}
xcreate\_adjsrc\_amplitude~t1~t2~ifile{[}0-5{]}~E/N/Z-ascii-files~{[}baz{]}.
\end{lyxcode}
\end{itemize}
\item \textbf{Run the kernel simulation}

>>>>

\begin{enumerate}
\item First, run a regular forward simlation (\texttt{SIMULATION\_TYPE =
1} and \texttt{SAVE\_FORWARD = .false.}). You can automatically set
these two variables using the \texttt{\small utils/change\_simulation\_type.pl}{\small{}
script:}{\small \par}
\begin{verbatim}
utils/change_simulation_type.pl -f
\end{verbatim}

@@

\begin{verbatim}
xcreate_adjsrc_traveltime t1 t2 ifile[0-5] E/N/Z-ascii-files [baz]
\end{verbatim}
where \texttt{t1} and \texttt{t2} are the start and end time of the
portion you are interested in, \texttt{ifile} denotes the component
of the seismograms to be used (0 for all three components, 1 for East,
2 for North, and 3 for vertical, 4 for transverse, and 5 for radial
component), \texttt{E/N/Z-ascii-files} indicate the three-component
displacement seismograms in the right order, and \texttt{baz} is the
back-azimuth of the station. Note that \texttt{baz} is only supplied
when \texttt{ifile} = 4 or 5.

\item Similarly, a sample program to compute adjoint sources for amplitude
finite-frequency kernels may be found in \texttt{utils/adjoint\_sources/amplitude}
and used in the same way as described for traveltime measurements
\begin{verbatim}
xcreate_adjsrc_amplitude t1 t2 ifile[0-5] E/N/Z-ascii-files [baz]
\end{verbatim}
\end{itemize}
\item \textbf{Run the kernel simulation}

### 2015-05-21 doc/USER_MANUAL/C_troubleshooting.tex

Please make sure that you have a working installation of a Fortran
compiler, a C compiler and an MPI implementation. You should be able
to compile this little program code:
\begin{lyxcode}
program~main~~~\\
~include~'mpif.h'~~~\\
~integer,~parameter~::~CUSTOM\_MPI\_TYPE~=~MPI\_REAL~~~\\
~integer~ier~~~\\
~call~MPI\_INIT(ier)~~~\\
~call~MPI\_BARRIER(MPI\_COMM\_WORLD,ier)~~~\\
~call~MPI\_FINALIZE(ier)~~~\\
~end
\end{lyxcode}
\item [{compilation fails stating:}] ~\\

\begin{lyxcode}
...~~~\\
~obj/program\_generate\_databases.o:~In~function~`MAIN\_\_':~~~\\
~program\_generate\_databases.f90:(.text+0x14):~undefined~\\~~~reference~to~`\_gfortran\_set\_std'~~~\\
~...~~~\\

\end{lyxcode}

@@

\begin{lyxcode}
...~~\\
~histogram~of~skewness~(0.~good~-~1.~bad):~~~\\
~0.0000000E+00~-~5.0000001E-02~27648~81.81818~\%~~~\\
~5.0000001E-02~-~0.1000000~0~0.0000000E+00~\%~~~\\
~...~~~\\

\end{lyxcode}

The first line means that you have 27,648 elements with a skewness
value between 0 and 0.05 (which means the element is basically not
skewed, just plain regular hexahedral element). The total number of
elements you have in this mesh is (see in the \texttt{output\_mesher.txt}
file a bit further down):
\begin{lyxcode}
...~~\\
~total~number~of~elements~in~entire~mesh:~33792~~~\\
~...~~~\\

\end{lyxcode}

@@

The histogram lists for this mesh also some stronger skewed elements,
for example the worst ones belong to:
\begin{lyxcode}
...~~\\
~0.6000000~-~0.6500000~2048~6.060606~\%~~~\\
~...~~~\\

\end{lyxcode}

@@

In such cases, try to change your \texttt{DATA/Par\_file} and set
e.g.:
\begin{lyxcode}
SUPPRESS\_UTM\_PROJECTION~=~.false.~~~\\

\end{lyxcode}

to be the same in \texttt{Mesh\_Par\_file} and \texttt{Par\_file}.
This flag should be identical when using the in-house mesher \texttt{xmeshfem3D},
\texttt{xgenerate\_databases} and \texttt{xspecfem3D} together to
run simulations.

@@

\begin{lyxcode}
...~~~\\
~{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}~~~\\
~{*}{*}{*}~Verification~of~simulation~parameters~{*}{*}{*}~~~\\
~{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}{*}~~~\\
~...~~~\\
~{*}{*}{*}~Minimum~period~resolved~=~4.308774~~~\\
~{*}{*}{*}~Maximum~suggested~time~step~=~6.8863556E-02~~~\\
~...~~~\\
~...~~~\\

\end{lyxcode}

>>>>

Please make sure that you have a working installation of a Fortran
compiler, a C compiler and an MPI implementation. You should be able
to compile this little program code:

{\footnotesize
\begin{verbatim}
 program main
 include 'mpif.h'
 integer, parameter :: CUSTOM_MPI_TYPE = MPI_REAL
 integer ier
 call MPI_INIT(ier)
 call MPI_BARRIER(MPI_COMM_WORLD,ier)
 call MPI_FINALIZE(ier)
 end
\end{verbatim}
}

\item [{compilation fails stating:}] ~
{\footnotesize
\begin{verbatim}
...
 obj/program_generate_databases.o: In function `MAIN__':
 program_generate_databases.f90:(.text+0x14): undefined reference to `_gfortran_set_std'
...
\end{verbatim}
}

@@

{\footnotesize
\begin{verbatim}
...
 histogram of skewness (0. good - 1. bad):
 0.0000000E+00 - 5.0000001E-02 27648 81.81818 %
 5.0000001E-02 - 0.1000000 0 0.0000000E+00 %
...
\end{verbatim}
}

The first line means that you have 27,648 elements with a skewness
value between 0 and 0.05 (which means the element is basically not
skewed, just plain regular hexahedral element). The total number of
elements you have in this mesh is (see in the \texttt{output\_mesher.txt}
file a bit further down):

{\footnotesize
\begin{verbatim}
...
 total number of elements in entire mesh: 33792
...
\end{verbatim}
}

@@

The histogram lists for this mesh also some stronger skewed elements,
for example the worst ones belong to:

{\footnotesize
\begin{verbatim}
...
 0.6000000 - 0.6500000 2048 6.060606 %
...

\end{verbatim}
}

@@

In such cases, try to change your \texttt{DATA/Par\_file} and set
e.g.:
\begin{verbatim}
SUPPRESS_UTM_PROJECTION = .false.
\end{verbatim}
to be the same in \texttt{Mesh\_Par\_file} and \texttt{Par\_file}.
This flag should be identical when using the in-house mesher \texttt{xmeshfem3D},
\texttt{xgenerate\_databases} and \texttt{xspecfem3D} together to
run simulations.

@@

{\footnotesize
\begin{verbatim}
...
 *********************************************
 *** Verification of simulation parameters ***
 *********************************************
...
 *** Minimum period resolved = 4.308774
 *** Maximum suggested time step = 6.8863556E-02
...
\end{verbatim}
}
